```{r "intro"}
#| echo: false
intro <- tryCatch(
    {
        description <- packageDescription("scRNAseqWorkshop")
        pkg <- description$Package
        version <- description$Version
        authors <- eval(parse(text = description$Authors))
        license <- description$License
        glue::glue(
            "**Package:** {pkg}<br/>\n", 
            "**Authors:** {paste(format(authors, include = c('given', 'family', 'role')), collapse = ', ')}<br/>\n", 
            "**Compiled:** {as.character(Sys.Date())}<br/>\n", 
            "**Package version:** {version}<br/>\n", 
            "**R version:** <b style='color:red;'>{R.version.string}</b><br/>\n", 
            "**BioC version:** <b style='color:red;'>{BiocManager::version()}</b><br/>\n", 
            "**License:** {license}<br/>"
        )
    }, 
    error = function(e) {"Local preview"}
)
```

`r intro`

# Welcome {-}

This is the landing page of the workshop entitled *Single-cell RNA-seq analysis with R/Bioconductor*. 

## Overview

This course will introduce biologists and bioinformaticians to the 
field of single-cell RNA sequencing. 
We will cover a range of software and analysis workflows that extend 
over the spectrum from the best practices in the filtering scRNA-seq data  
to the downstream analysis of cell clusters and temporal ordering. 
This course will help the attendees gain accurate insights in pre-processing, 
analysis and interpretation of scRNA-seq data.

We will start by introducing general concepts about single-cell RNA-sequencing. 
From there, we will then continue to describe the main analysis steps to go from 
raw sequencing data to processed and usable data. We will present classical 
analysis workflows, their output and the possible paths to investigate for 
downstream analysis.  

Throughout this workshop, we will put an emphasis on `R/Bioconductor` ecosystem 
and the different packages which will be used to analyse datasets and learn new approaches. 

## Format

The course is structured in modules over five days.  

During the first 1/3 of the day, formal lectures will cover the key 
concepts required to understand the principles of scRNA-seq analysis (~2h).

Following these lectures, practical examples will be shown to illustrate how 
to translate the acquired knowledge into functional `R` code (~1h). At this stage, 
trainees will get acquainted with state-of-the-art packages for scRNAseq analysis 
as well as the best practices in bioinformatics. 

During the second half of the day (3h), trainees will work by themselves, following 
guided exercises to improve their understanding of scRNAseq analysis workflow.
Hints and solution are provided for each exercise. The exercises will mainly focus 
on specific concepts introducted earlier that day.
However, analytical steps studied throughout the previous days will also 
be integrated so that towards the end of the week, trainees are fully able 
to perform fundamental scRNAseq analyses from beginning to end. 

Office hours will take place during the last hour of the exercises. An 
instructor will be available to answer individual questions related to daily exercises. 
A Slack channel will also be available so that Q&A are available for everybody.

## Learning outcomes

At the end of this course, you should be able to:

- Understand the pros/cons of different single-cell RNA-seq methods
- Process and QC of scRNA-seq data
- Normalize scRNA-seq data
- Correct for batch effects
- Visualise the data and applying dimensionality reduction
- Perform cell clustering and annotation 
- Perform differential gene expression analysis
- Infer cell trajectory and pseudotime, and perform temporal differential expression 

## Instructor

[Dr. Jacques Serizay](mailto:jacques.serizay@pasteur.fr)

# Program {-}

## Day 1 {-}

- **\[1h\]** Lecture 1 - General introduction to single-cell RNA-seq experimental design
    [[Pptx]](Presentations/Lecture1_experimental-design.pptx)

- **\[1h\]** Lecture 2 - scRNAseq: from raw sequencing files to count matrix
    [[Pptx]](Presentations/Lecture2_bcl-to-matrix.pptx)

- **\[1h\]** Demonstration 1 - From fastq to count matrix
    [[HTML](Demonstration/Day1_demonstration.qmd) | [qmd (save to open in RStudio)](Demonstration/Day1_demonstration.qmd)]

-  **\[3h\]** Homework - From bcl to count matrix
    [[HTML](Exercises/Day1_exercises.qmd) | [qmd (save to open in RStudio)](Exercises/Day1_exercises.qmd)]

## Day 2 {-}

- **\[1h\]** Lecture 3 - Filtering cells in droplet-based scRNAseq data
    [[Pptx]](Presentations/Lecture3_doublet-cells.pptx)

- **\[1h\]** Lecture 4 - Normalizing scRNAseq data
    [[Pptx]](Presentations/Lecture4_normalization.pptx)

- **\[BONUS\]** R/Bioconductor essentials: `GRanges` and `*Experiment` classes [[Pptx]](docs/R-essentials.pptx)

- **\[1h\]** Demonstration 2 - Leveraging R/Bioconductor for single-cell analyses
    [[HTML](Demonstration/Day2_demonstration.qmd) | [qmd (save to open in RStudio)](Demonstration/Day2_demonstration.qmd)]

-  **\[3h\]** Homework - scRNAseq analysis with R/Bioconductor (1/3)
    [[HTML](Exercises/Day2_exercises.qmd) | [qmd (save to open in RStudio)](Exercises/Day2_exercises.qmd)]

## Day 3 {-}

- **\[2h\]** Lecture 5 - Clustering cells in scRNAseq 
    [[Pptx]](Presentations/Lecture5_clustering.pptx)

- **\[1h\]** Demonstration 3 - Dimensional reduction visualization and clustering
    [[HTML](Demonstration/Day3_demonstration.qmd) | [qmd (save to open in RStudio)](Demonstration/Day3_demonstration.qmd)]

-  **\[3h\]** Homework - scRNAseq analysis with R/Bioconductor (2/3)
    [[HTML](Exercises/Day3_exercises.qmd) | [qmd (save to open in RStudio)](Exercises/Day3_exercises.qmd)]

## Day 4 {-}

- **\[1h\]** Lecture 6 - Cell type annotations 
    [[Pptx]](Presentations/Lecture6_annotations.pptx)

- **\[1h\]** Lecture 7 - Batch correction 
    [[PDF]](Presentations/Lecture7_batchcorrection.pdf)

- **\[1h\]** Demonstration 4 - Annotation transfer with `scmap`
    [[HTML](Demonstration/Day4_demonstration.qmd) | [qmd (save to open in RStudio)](Demonstration/Day4_demonstration.qmd)]

-  **\[3h\]** Homework - scRNAseq analysis with R/Bioconductor (3/3)
    [[HTML](Exercises/Day4_exercises.qmd) | [qmd (save to open in RStudio)](Exercises/Day4_exercises.qmd)]

## Day 5 {-}

- **\[2h\]** Lecture 8 - Trajectory inference and RNA velocity
    [[Pptx]](Presentations/Lecture8_pseudotime.pptx)

- **\[1h\]** Demonstration 5 - Trajectory inference in multiciliated cells 
    [[HTML](Demonstration/Day5_demonstration.qmd) | [qmd (save to open in RStudio)](Demonstration/Day5_demonstration.qmd)]

-  **\[3h\]** Homework - Advanced scRNAseq topics: trajectory inference and RNA velocity
    [[HTML](Exercises/Day5_exercises.qmd) | [qmd (save to open in RStudio)](Exercises/Day5_exercises.qmd)]


# Docker image {-}

A `Docker` image built from this repository is available here: 

👉 [ghcr.io/js2264/scrnaseqworkshop](https://ghcr.io/js2264/scrnaseqworkshop) 🐳

::: {.callout-tip icon='true'}
## Get started now 🎉

You can get access to all the packages used in this book in < 1 minute, 
using this command in a terminal: 

```{sh "docker", filename="bash"}
#| eval: false
docker run -it ghcr.io/js2264/scrnaseqworkshop:devel R
```

:::

# RStudio Server {-}

An RStudio Server instance can be initiated from the `Docker` image as follows: 

```{sh "rstudio", filename="bash"}
#| eval: false
docker run \
    --volume <local_folder>:<destination_folder> \
    -e PASSWORD=OHCA \
    -p 8787:8787 \
    ghcr.io/js2264/scrnaseqworkshop:devel
```

The initiated RStudio Server instance will be available at 
[https://localhost:8787](https://localhost:8787).

# Course material {-} 

The workshop content is available [at this adress: https://github.com/js2264/scRNAseq-workshop](https://github.com/js2264/scRNAseq-workshop). 

You can clone it locally with `git`: 

```sh
git clone https://github.com/js2264/scRNAseq-workshop.git
```

To download it without the command-line tool `git`, go to 
[the GitHub repo page](https://github.com/js2264/scRNAseq-workshop), click on the 
green `Code` button, then `Download ZIP`. Beware, the download may take a significant 
time based on your internet connection (several hundreds MB). 


# Session info {-}

::: {.callout-note collapse="true"}

## Click to expand 👇

```{r "session info"}
#| cache: false
sessioninfo::session_info(
    installed.packages()[,"Package"], 
    include_base = TRUE
)
```

:::
